---
title: "MCC NSE CHA"
author: "Kenneth Daily"
date: "06/04/2014"
output: html_document
---


```{r libs, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(reshape)
library(plyr)
library(RColorBrewer)
library(knitr)
library(xtable)
library(gridExtra)

options(xtable.type="html", xtable.caption.placement="top", xtable.include.row.names=FALSE)
opts_chunk$set(comment=NA, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.width=10, fig.height=5)

lab.thresh.NSE <- 17
lab.thresh.ChrAnew <- 84.7

findclosestdatebefore <- function(dates, refdate, checkorder=FALSE) {
  
  if (all(is.na(dates))) {
    return(NA)
  }
  
  if (is.na(refdate)) {
    return(NA)
  }
  
  # find the closest date occuring before a reference date
  result <- dates[with(data.pt, which.min(dates - refdate))]
  if (checkorder & result > refdate) {
    result <- NA
    }
  
  result
}


ddpfcl <- function(a) {
  refdate <- a$firstprg[1]
  findclosestdatebefore(a$Date, refdate)
}

```

```{r loaddata}

data <- read.csv("data/2014-06-03_NSE_ChrA_all.csv",
                 colClasses=c("factor", "character", "numeric", "numeric", "numeric"))

metadata <- read.csv("data/2014-06-03_MCC_Dx_Px_dates.csv")

metadata.cols <- c("id", "mccdxdat", "progfree", "firstprg", "nseprgda", "nse1date", "chrne1da", "chraprgdanew", "chraprgdaold")
metadata.use <- metadata[, metadata.cols]
metadata.use[metadata.use == ""] <- NA

data.merge <- merge(data, metadata.use, by.x="PtNum", by.y="id")

data.merge <- transform(data.merge,
                        Date=as.Date(Date, format="%m/%d/%Y"),
                        mccdxdat=as.Date(mccdxdat, format="%m/%d/%Y"),
                        progfree=as.Date(progfree, format="%m/%d/%Y"),
                        firstprg=as.Date(firstprg, format="%m/%d/%Y"),
                        nse1date=as.Date(nse1date, format="%m/%d/%Y"),
                        nseprgda=as.Date(nseprgda, format="%m/%d/%Y"),
                        chrne1da=as.Date(chrne1da, format="%m/%d/%Y"),
                        chraprgdanew=as.Date(chraprgdanew, format="%m/%d/%Y"),
                        chraprgdaold=as.Date(chraprgdaold, format="%m/%d/%Y")
                        )

data.merge$chraprgda <-data.merge$chraprgdanew
baddate <- is.na(data.merge$chraprgdanew)
data.merge$chraprgda[baddate] <- data.merge$chraprgdaold[baddate]

data.merge <- transform(data.merge,
                        DateDays=difftime(Date, mccdxdat, units="days"),
                        
                        progfreedays=ifelse(is.na(progfree),
                                            difftime(firstprg, mccdxdat, units="days"),
                                            difftime(progfree, mccdxdat, unit="days")),
                        
                        firstprgdays=difftime(firstprg, mccdxdat, units="days"),
                        
                        DateWeeks=difftime(Date, mccdxdat, units="weeks"),

                        progfreeweeks=ifelse(is.na(progfree),
                                             difftime(firstprg, mccdxdat, units="weeks"),
                                             difftime(progfree, mccdxdat, unit="weeks")),
                        
                        progressed=factor(is.na(firstprg),
                                           levels=c(TRUE, FALSE),
                                           labels=c("No Progression", "Progression"),
                                           ordered=TRUE),
                        
                        closestnsedate=Date == nseprgda,
                        closestchradate=Date == chraprgda
                        )

beforeprogression <- with(data.merge, Date <= firstprg)
beforeprogression[is.na(beforeprogression)] <- "No Progression"

data.merge$beforeprogression <- factor(beforeprogression,
                                       levels=c("No Progression", TRUE, FALSE),
                                       labels=c("No Progression", "Before Progression", "After Progression"),
                                       ordered=TRUE)

chra.old.cutpoints <- c(0, 25, 53.75, 1000)
chra.new.cutpoints <- c(0, 84.7, 105.65, 1000)

data.merge <- transform(data.merge,
                        ChrAo.cut=cut(ChrAo, breaks=chra.old.cutpoints, 
                                      labels=c("WNL", "above normal", "high"),
                                      ordered_result=TRUE),
                        ChrAn.cut=cut(ChrAn, breaks=chra.new.cutpoints,
                                      labels=c("WNL", "above normal", "high"),
                                      ordered_result=TRUE)
                        )

data.merge$ChrA.cut <- factor(with(data.merge, ifelse(is.na(ChrAn.cut), as.character(ChrAo.cut), as.character(ChrAn.cut))),
                              levels=c("WNL", "above normal", "high"), ordered=TRUE)

# foo <- ddply(data.merge, .(PtNum), ddpfcl)
# colnames(foo) <- c("PtNum", "mintestdate")
# bar <- merge(data.merge, foo, by="PtNum")
# head(data.merge[, c("beforeprogression", "progressed", "progfree", "firstprg", "firstprgdays")])

```

## NSE
- WNL: < 17 ug
- above normal: >= 17 until 24.3
- high: > 24.3

## Summary of data above lab-determined NSE threshold of `r lab.thresh.NSE`
```{r}
summary(subset(data.merge, NSE > lab.thresh.NSE)$NSE)
```

### Table of progression status and NSE data availability
```{r, results='asis'}
foo <- ddply(data.merge, .(PtNum), function(x) data.frame(progressed=any(x$progressed == "Progression"), NSE=any(!is.na(x$NSE))))
tbl <- with(foo, table(progressed, NSE))
rownames(tbl) <- c("No Progression", "Progression")
colnames(tbl) <- c("NSE not available", "NSE available")
xtable(tbl)
```

### NSE summary distribution, testing date progression status

```{r nsesummary, include=FALSE}
nse.cutpoints <- data.frame(y=c(17, 24.3))

pointspersample <- ddply(data.merge, .(PtNum), nrow)
samples.keep <- subset(pointspersample, V1 > 3)$PtNum
# samples.keep <- c(18, 31, 38, 53, 64, 8)

plot.data <- subset(data.merge, !is.na(NSE) & as.numeric(DateDays) >= 0 & PtNum %in% samples.keep)

p1 <- ggplot(plot.data, aes(x=NSE, group=beforeprogression, color=beforeprogression))
p1 <-p1 + geom_density()
p1 <- p1 + scale_color_manual(values=c("red", "blue", "green"))
p1 <- p1 + theme_bw()
p1 <- p1 + theme(axis.text.y=element_blank(), axis.title.y=element_blank(),
                 axis.title.x=element_blank(), legend.position="none")

p2 <- ggplot(plot.data, aes(y=NSE, x=beforeprogression, color=beforeprogression))
p2 <-p2 + geom_boxplot()
p2 <- p2 + scale_color_manual(values=c("red", "blue", "green"))
p2 <- p2 + coord_flip()

p2 <- p2 + theme_bw()
p2 <- p2 + theme(axis.text.y=element_blank(), axis.title.y=element_blank(), legend.position="bottom")

grid.arrange(p1, p2, nrow=2, heights = unit(c(2,1), "null"))
```

### NSE data per patient

```{r nseplot1}
nse.cutpoints <- data.frame(y=c(17, 24.3))

# pointspersample <- ddply(data.merge, .(PtNum), nrow)
# samples.keep <- subset(pointspersample, V1 > 3)$PtNum
# samples.keep <- c(18, 31, 38, 53, 64, 8)
samples.keep <- metadata$id

plot.data <- subset(data.merge, !is.na(NSE) & as.numeric(DateDays) >= 0 & PtNum %in% samples.keep)

p <- ggplot(plot.data, aes(x=as.numeric(DateDays), y=NSE, group=PtNum))
p <- p + geom_line(aes(color=PtNum), size=1)
p <- p + geom_point(aes(color=PtNum), size=3)
# p <- p + facet_wrap(~ PtNum)

# Get the first point to highlight
min.plot.data <- ddply(plot.data, .(PtNum), function(x) x[which(min(x$Date) == x$Date), ])
p <- p + geom_point(data=min.plot.data, aes(x=as.numeric(DateDays), y=NSE, group=PtNum), shape="*", size=8)

# get the closest test date to progression
closest.prog <- subset(plot.data, closestnsedate & progressed == "Progression")
p <- p + geom_point(data=closest.prog, aes(x=as.numeric(DateDays), y=NSE, group=PtNum), shape=8, size=6)

# Draw cutpoints
p <- p + geom_hline(data=nse.cutpoints, aes(yintercept=y))

p <- p + facet_grid(~ progressed)

p <- p + labs(x="Days from Dx", y="NSE (ug)")
p <- p + theme_bw()
p <- p + theme(legend.position="none")
print(p)
```

```{r nseplotfxn}

plot.nse <- function(plot.data, nsecutpoints, title=NA, ...) {
p <- ggplot(plot.data, aes(x=as.numeric(DateDays), y=NSE, group=PtNum))
p <- p + geom_line(size=1)
p <- p + geom_point( size=3)

# mark the date of progression, if it exists
prog.data <- unique(subset(plot.data, !is.na(firstprgdays))[, c("PtNum", "firstprgdays")])
p <- p + geom_vline(data=prog.data, aes(xintercept=as.numeric(firstprgdays)), color="red")
# p <- p + geom_point(data=min.plot.data, aes(x=as.numeric(DateDays), y=NSE, group=PtNum), shape="*", size=8, color="red")

# # get the closest test date to progression
# closest.prog <- subset(plot.data, closestnsedate & progressed == "Progression")
# p <- p + geom_point(data=closest.prog, aes(x=as.numeric(DateDays), y=NSE, group=PtNum), shape=8, size=6)

# Draw cutpoints
p <- p + geom_hline(data=nse.cutpoints, aes(yintercept=y))

p <- p + facet_wrap(~ PtNum, scale="free_x")
p <- p + labs(x="Days from Dx", y="NSE (ug)", title=title)

p <- p + theme_bw()
p <- p + theme(legend.position="none")

p

}

```


```{r nseplot2, fig.width=15, fig.height=7}
# pointspersample <- ddply(data.merge, .(PtNum), nrow)
# samples.keep <- subset(pointspersample, V1 > 3)$PtNum
# samples.keep <- c(18, 31, 38, 53, 64, 8)
samples.keep <- metadata$id
plot.data <- subset(data.merge, !is.na(NSE) & as.numeric(DateDays) >= 0 & PtNum %in% samples.keep)

p.progressed <- plot.nse(subset(plot.data, progressed == "Progression"), nse.cutpoints, title="Progression")
p.noprogressed <- plot.nse(subset(plot.data, progressed == "No Progression"), nse.cutpoints, title="No Progression")

grid.arrange(p.noprogressed, p.progressed, ncol=2)

```

## ChrA
### ChrA old

- WNL: < 25 U/ml
- above normal: >= 25 until < 53.75
- high: >= 53.75

### ChrA new
- WNL: < 84.7 ng/ml
- above normal: >= 84.7 until < 105.65
- high: >= 105.65

### Summary of data above lab-determined new ChrA threshold of `r lab.thresh.ChrAnew`
```{r}
summary(subset(data.merge, ChrAn >= lab.thresh.ChrAnew)$ChrAn)
```


### ChrA data per patient

```{r chraplot1}
plot.data <- subset(data.merge, !is.na(ChrA.cut))

p <- ggplot(plot.data, aes(x=as.numeric(DateDays), y=ChrA.cut, group=PtNum))
p <- p + geom_line(aes(color=PtNum), size=1)
p <- p + geom_point(aes(color=PtNum), size=3)
# p <- p + facet_wrap(~ PtNum)

# Get the first point to highlight
min.plot.data <- ddply(plot.data, .(PtNum), function(x) x[which(min(x$Date) == x$Date), ])
p <- p + geom_point(data=min.plot.data, aes(x=as.numeric(DateDays), y=ChrA.cut, group=PtNum), shape="*", size=8)

# # get the closest test date to progression
# closest.prog <- subset(plot.data, closestchradate & progressed == "Progression")
# p <- p + geom_point(data=closest.prog, aes(x=as.numeric(DateDays), y=ChrA.cut, group=PtNum), shape=8, size=6)

p <- p + facet_grid(~ progressed)

p <- p + labs(x="Days from Dx", y="ChrA level")
p <- p + theme_bw()
p <- p + theme(legend.position="none")
print(p)
```

```{r chraplotfxn}

plot.chra <- function(plot.data, title=NA, ...) {
p <- ggplot(plot.data, aes(x=as.numeric(DateDays), y=ChrA.cut, group=PtNum))
p <- p + geom_line(size=1)
p <- p + geom_point( size=3)

# mark the date of progression, if it exists
prog.data <- unique(subset(plot.data, !is.na(firstprgdays))[, c("PtNum", "firstprgdays")])
p <- p + geom_vline(data=prog.data, aes(xintercept=as.numeric(firstprgdays)), color="red")
# p <- p + geom_point(data=min.plot.data, aes(x=as.numeric(DateDays), y=NSE, group=PtNum), shape="*", size=8, color="red")

# # get the closest test date to progression
# closest.prog <- subset(plot.data, closestchradate & progressed == "Progression")
# p <- p + geom_point(data=closest.prog, aes(x=as.numeric(DateDays), y=ChrA.cut, group=PtNum), shape=8, size=6)

p <- p + facet_wrap(~ PtNum, scale="free_x")
p <- p + labs(x="Days from Dx", y="ChrA level", title=title)

p <- p + theme_bw()
p <- p + theme(legend.position="none")

p

}

```

```{r plotchra2, fig.width=15, fig.height=7}
plot.data <- subset(data.merge, !is.na(ChrA.cut))

p.progressed <- plot.chra(subset(plot.data, progressed == "Progression"), title="Progression")
p.noprogressed <- plot.chra(subset(plot.data, progressed == "No Progression"), title="No Progression")

grid.arrange(p.noprogressed, p.progressed, ncol=2)


```
